%Options
% FhD -> Changes the template for Univesity of dormtund
% HSHL -> Changes the template for Univesity of Lippstadt
% These change the type accordingly
% ProjectWork  
% BachelorThesis 
% MasterThesis 
% ResearchThesis 

\documentclass[FhD, ResearchThesis]{teamhenkler}
\title{A Comparative Analysis of System Architectures for Collaborative SLAM}
\author{Aditya Kumar}
\supervisor{Prof. Dr. Stefan Henkler}
\cosupervisor{-}
\graddate{\today}
\matriculationnumber{7219675}


%Change this if you are doing it with a company
%\companyname{Company Name}
%\companylogo{\logopath} %replace \logopath with the company logo, comment it out if you dont want any logo

\begin{document}

\maketitle

\begin{abstract}
Collaborative Simultaneous Localisation and Mapping (C-SLAM) enables teams of robots to construct a shared environment map while estimating their trajectories. As multi-robot systems move toward larger deployments in dynamic settings, the backend architecture becomes a primary determinant of end-to-end performance. However, experimentally controlled comparisons between different C-SLAM architectures with similar frontend-backend processing under realistic network effects remain scarce. 

This research thesis presents a systematic, implementation-grounded comparison of two representative backend paradigms within a unified evaluation framework: a centralised incremental pose graph optimiser and a decentralised DDF-style distributed optimiser. To ensure architectural fairness, both backends consume identical pre-computed factor streams from the COSMO-Bench dataset and are evaluated using a fully automated ROS 2 testbed that applies packet loss, transmission delay, and bandwidth limitations in a controlled manner. 

Performance is quantified using localisation accuracy, latency and convergence behaviour, communication bandwidth, and computational resource utilisation. The centralised backend achieves superior global consistency (ATE RMSE 1.89–2.53 m across evaluated scenarios) and rapid team-level convergence $(T(team)\_conv < 0.2 s)$ as team size grows from three to five robots. However, it incurs asymmetric communication patterns with downlink traffic comprising 80–90\% of total bandwidth (26–49 MiB per run) and concentrates processing at the server. 

The decentralised backend demonstrates regime-dependent performance characteristics. Under ProRadio networking, it maintains competitive accuracy (ATE RMSE 2.72–3.82 m) with peer-to-peer traffic 12–45\% lower than centralised uplink requirements. Under WiFi networking, accuracy degrades at larger team sizes (ATE RMSE 6.66 m at five robots) despite increased interface coordination activity. Both architectures exhibit sensitivity to severe bandwidth constraints (0.25 Mbps caps increase dissemination latency 8–10× for centralised; interface stabilisation becomes inconsistent for decentralised), but the decentralised implementation encounters stability failures under intermittent connectivity that limit robustness claims in the current implementation.

Overall, neither centralised nor decentralised C-SLAM paradigms are universally optimal; their effectiveness depends on specific deployment requirements. Centralised architectures are preferable when stable infrastructure is available, global consistency is critical, and team sizes remain small to medium ( $<= 5 $ robots in the evaluated scenarios). Decentralised architectures are appropriate when bandwidth is severely constrained, peer-to-peer communication is reliable, and the constraint regime supports effective distributed coordination—though practical deployment requires resolving the identified software stability issues under network intermittency.
\end{abstract}


\makedeclaration
\maketoc

\input{src/introduction}
\input{src/fundamentals}
\input{src/state_art}
\input{src/methodology}
\input{src/implementation}
\input{src/evaluation}
%\input{src/rough}
\input{src/conclusion}
\bibliographystyle{plain}
\bibliography{src/bib} 
\clearpage

\appendix

\input{src/A1_Appendix}

\end{document}
